PACKER: An Exemplar Model of Category Generation

### 2017-01-22 Manuscripts/cogsci-2017/paper.tex
update to proceedings, figs

@@

The work presented by Ward \& Colleagues, as well as that of \citet{jern2013probabilistic}, undoubtedly captures a core principle of the creative use of conceptual knowledge: category generation is informed by the distributional characteristics of known concepts in the target domain. However, this work fails to describe the inverse: how do generated categories \textit{differ} from known categories? In this paper, we introduce a novel exemplar-based approach to category generation, the PACKER model (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates new categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity).

% Is this evident?: "we are not suggesting Jern \& Kemp have failed to capture a core aspect of generation, but that there are other properties that can be explained through one of the field's oldest traditions."

@@

@@

\subsubsection{Summary.}

@@

\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example-prob-spaces.pgf}
    \caption{Predicted generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. Areas in which generation is not likely are shaded white; high probability areas are shaded blue. \textit{Left}: Predictions given $\{\phi = -1$, $\gamma = 0\}$ (contrast influence only). \textit{Center}: Predictions given $\{\phi = 0$, $\gamma = 1\}$ (target influence only).  \textit{Right}: Predictions given $\{\phi = -1$, $\gamma = 1\}$ (both constraints considered).  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

Although PACKER appeals to the well-studied exemplar framework, it remains unique as an account of conceptual generation because, unlike the accounts put forward by \citet{jern2013probabilistic} and Ward \& colleagues, it specifies precisely how generated categories should \textit{differ} from known classes in the domain. In this paper, we explore whether an exemplar-similarity approach can offer insight into the creative use of conceptual knowledge. In particular, we focus on a key difference between PACKER and previous accounts: unlike the model proposed by \citeauthor{jern2013probabilistic}, the distributional properties of known categories affects performance in PACKER only insofar as it constrains inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories.

@@

The behavioral experiment described below was designed to test the prediction made that the \textit{location} of contrast categories (as opposed to their structure), influences generation. As such, our experiment manipulates only the location of the known categories within the domain (the structure of the categories is held constant).

The experiment follows the paradigm set by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. See Figure \ref{fig:middle-bottom-conditions} for a depiction of the experimental conditions. Our manipulation is fairly `weak': the conditions have only a slight difference in the Y-Axis position of the Alpha category. In the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

@@

Because the distributional characteristics of the domain do not differ between the Middle and Bottom conditions, the \citet{jern2013probabilistic} model predicts that there should be no difference in the categories generated within each domain. However, the PACKER model proposes that exemplars are generated such that each category has strong within-class similarity and weak between-class similarity. By consequence, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle, as stimuli in this area possess weaker similarity to members of the Alpha category. Conversely, the bottom areas of the space should be less frequently used for generation in the Bottom condition compared to Middle, as stimuli in this area possess greater similarity to the Alpha category.

@@

Participants began the experiment with a short training phase (3 blocks of 4 trials), where they iteratively observed exemplars belonging to the `Alpha' category. Participants were instructed to learn as much as they can about the Alpha category, and that they would answer a series of test questions afterwards. On each trial, a single exemplar from the Alpha category was presented, and participants were given as much time as they desired before moving on.

@@

As in the \citet{jern2013probabilistic} experiments, generation was completed using a sliding-scale interface. Participants were presented with two scales that could be used to change the color and size of the generated example. An on-screen preview of the example updated whenever one of the features was changed. Participants could generate any example along an evenly-spaced 9x9 grid, and they could change the example as much as they wanted before moving on. Neither the members of the Alpha category nor the previously generated Beta examples were visible during generation.

@@

Because the conditions differ only in their location along the vertical axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in the generation of Beta examples above and below the contrast category. A highly detailed picture of these data is shown in Figure \ref{fig:middle-bottom-yranges}, wherein each participant's generated category is depicted in terms of its use of the vertical axis. As is evident in Figure \ref{fig:middle-bottom-yranges}, we observed a great deal of variance in generation strategy: whereas some participants generated all four Beta examples within a narrow range of the Y axis, others generated Beta examples along a fairly wide range.

@@

To obtain a more abstracted view of our results, we determined the number of participants in each condition who placed at least one Beta example within the top `rows' of the space (one of the top two possible Y axis values), as well as the number of participants who placed at least one example in one of the bottom rows. The resulting contingency tables data are shown in Table \ref{table:subset-table}.

@@

Fisher's Exact Tests reveal that a greater number of participants assigned to the Middle condition generated at least one Beta example in the bottom rows of the space, $p < 0.001$. Participants in the Bottom condition were more marginally more likely to generate at least one Beta example in the top rows of the space, $p = 0.088$. The conditions did not differ in the number of participants who placed at least one Beta example in the top rows \textit{and} at least one Beta example in the bottom rows, $p = 0.14$. Finally, a greater number of Middle participants generated all four Beta examples in the bottom rows of the space, $p = 0.003$, and a greater number of Bottom participants generated all four examples in the top rows of the space, $p = 0.017$.

@@

\section{Simulations}
\section{Discussion}
\section{Acknowledgments}
Support for this research was provided by the Office of the Vice Chancellor for Research and Graduate Education at the University of Wisconsin - Madison with funding from the Wisconsin Alumni Research Foundation. We would like to thank Kenneth J. Kurtz for helpful comments on this work.
% any other thanks?

>>>>

The work presented by Ward \& Colleagues, as well as that of \citet{jern2013probabilistic}, undoubtedly captures a core principle of the creative use of conceptual knowledge: category generation is informed by the distributional characteristics of known concepts in the target domain. However, this work fails to describe the inverse: how do generated categories \textit{differ} from known categories? In this paper, we introduce a novel exemplar-based approach to category generation, the PACKER model (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates new categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity). We then report a behavioral experiment testing the predictions of the PACKER model, and report formal simulations comparing PACKER to the model proposed by \cite{jern2013probabilistic}.

@@

\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example-prob-spaces.pgf}
    \caption{PACKER generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. Areas in which generation is not likely are shaded white; high probability areas are shaded blue. \textit{Left}: Predictions given $\{\phi = -1$, $\gamma = 0\}$ (contrast influence only). \textit{Center}: Predictions given $\{\phi = 0$, $\gamma = 1\}$ (target influence only).  \textit{Right}: Predictions given $\{\phi = -1$, $\gamma = 1\}$ (both constraints considered).  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

@@

% Note something about euclidean distance for integral dims?  Or is the field over that?

@@

@@

In the sections below, we report a behavioral experiment testing a key difference between PACKER and previous accounts: unlike the model proposed by \citeauthor{jern2013probabilistic}, the distributional properties of known categories affects performance in PACKER only insofar as it constrains inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories. We then report formal simulations evaluating the fit of the PACKER and \citet{jern2013probabilistic} models to our data.

@@

The behavioral experiment described below was designed to test whether the \textit{location} of contrast categories (as opposed to their structure), influences generation. As such, our experiment manipulates only the location of the known categories within the domain (the structure of the categories is held constant).

@@

The experiment follows the paradigm set by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. Our manipulation is fairly `weak' (see Figure \ref{fig:middle-bottom-conditions}): the conditions have only a slight difference in the Y-Axis position of the Alpha category. In the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

Because the distributional characteristics of the domain do not differ between the Middle and Bottom conditions, the \citet{jern2013probabilistic} model predicts that there should be no difference in the categories generated within each domain. However, the PACKER model proposes that exemplars are generated such that each category has strong within-class similarity and weak between-class similarity. By consequence, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle, as stimuli in this area possess weaker similarity to members of the Alpha category. Conversely, the bottom areas of the space should be less frequently used for generation in the Bottom condition compared to Middle, as stimuli in this area possess greater similarity to the Alpha category. 
% there has GOT to be a more concise way of expressing those final two sentences...

@@

Participants began the experiment with a short training phase (3 blocks of 4 trials), where they iteratively observed exemplars belonging to the `Alpha' category. Participants were instructed to learn as much as they can about the Alpha category, and that they would answer a series of test questions afterwards. On each trial, a single exemplar from the Alpha category was presented, and participants were given as much time as they desired before moving on. Participants were shown a preview of the range of possible colors and sizes prior to training.

@@

As in the \citet{jern2013probabilistic} experiments, generation was completed using a sliding-scale interface. Participants were presented with two scales that could be used to change the color and size of the generated example. An on-screen preview of the example updated whenever one of the features was changed. Participants could generate any example along an evenly-spaced 9x9 grid, and they could change the example as much as they wanted before moving on. Neither the members of the Alpha category nor the previously generated Beta examples were visible during generation. Participants were not informed if they generated a member of the Alpha category, but could not complete the trial if they generated the same Beta two times.

@@

Because the conditions differ only in their location along the vertical axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in the generation of Beta examples above and below the contrast category. A highly detailed depiction of our results is shown in Figure \ref{fig:middle-bottom-yranges}, wherein each participant's generated category is depicted in terms of its use of the vertical axis. As is evident in Figure \ref{fig:middle-bottom-yranges}, we observed a great deal of individual differences in generation strategy: whereas some participants generated all four Beta examples within a narrow range of the Y axis, others generated Beta examples along a fairly wide range.

@@

To more specifically evaluate the key predictions made by PACKER, we determined the number of participants in each condition who placed at least one Beta example within the top `rows' of the space (one of the top two possible Y axis values), as well as the number of participants who placed at least one example in one of the bottom rows. The resulting contingency tables data are shown in Table \ref{table:subset-table}.

@@

Fisher's Exact Tests reveal that a greater number of participants assigned to the Middle condition generated a Beta example in the bottom rows of the space, $p < 0.001$. Participants in the Bottom condition were more marginally more likely to generate a Beta example in the top rows of the space, $p = 0.088$. The conditions did not differ in the number of participants who placed Beta examples in the top rows \textit{and} the bottom rows, $p = 0.14$. Finally, a greater number of Middle participants generated all four Beta examples in the bottom rows of the space, $p = 0.003$, and a greater number of Bottom participants generated all four examples in the top rows of the space, $p = 0.017$.

These results are broadly supportive of PACKER's predictions, and are not predicted by existing accounts \citep[i.e.,][]{jern2013probabilistic}. However, the results described above are somewhat commonsense: they simply demonstrate that Beta examples tend to be generated in areas not occupied by the Alphas. Beyond these effects, we observed a great deal of differences in generation strategy. Figure \ref{fig:range-diff-gradient} depicts the average difference in range between the two features (e.g., horizontal -- vertical) across the participant-generated categories, with respect to the location of each category's members. These data reveal highly systematic patterns of generation: whereas many participants generated categories with more vertical range (i.e., `column'-like categories), others generated categories with horizontal range (`row'-like categories). Furthermore, these two different types of categories appear in distinct locations of the domain space. Whereas vertically aligned categories more often appear to the left or right of the Alpha class, horizontally aligned categories appear above and below the Alpha class. Thus, even beyond generating Betas in locations not occupied by the Alphas, participants appear to modify the internal structure of their categories in order to maximize distance from the Alphas.

\begin{figure}[ht!]
    \begin{center}
    \inputpgf{figs/}{range-diff-gradient.pgf}
    \caption{Generated category structure as a function of location in the domain space. Orange areas in each gradient correspond to stimuli that were commonly generated into category possessing greater vertical (Y-Axis) range. Purple areas correspond to categories possessing greater horizontal range. White areas correspond to equal range along both features (or infrequent generation).}
    \label{fig:range-diff-gradient}
    \end{center}
\end{figure}

@@

Our behavioral experiment revealed strong effects of the role of category location on generation. Not only are participants more likely to generate Betas in locations

\section{Simulations}

We conducted formal simulations to our data using the PACKER model described above, as well as a conjugate implementation of the hierarchical sampling model proposed by \citet{jern2013probabilistic}.

% do we need to describe the implementation?

\subsection{Model-Fitting}
In order to obtain an overall sense of each model's ability to explain our results, we fitted each model trial-wise, maximizing the log-likelihood of the its predictions against our results. For each trial, models were initialized with the participant's Alpha category (Middle or Bottom), and the configuration of the participant's Beta category during that trial. Four parameters were fitted for the PACKER model: $c$, $\phi$, $\gamma$, and $\theta$. Likewise, four parameters were fitted to the \cite{jern2013probabilistic} model: $\kappa$, $\gamma$, $\nu$, and $\theta$. Parameters were not allowed to vary between participants or conditions -- the goal was to obtain the best-fit to our entire dataset.

\section{Discussion}

The creative use of conceptual knowledge is a highly intriguing yet understudied topic in category learning research. In this paper, we presented a novel exemplar-based approach to explaining category generation.

\section{Acknowledgments}
Support for this research was provided by the Office of the Vice Chancellor for Research and Graduate Education at the University of Wisconsin - Madison with funding from the Wisconsin Alumni Research Foundation. We would like to thank Kenneth Kurtz for helpful comments on this work. We also thank Alan Jern \& Charles Kemp for providing code and data from their experiments.
% any other thanks?
